---
title: "Putting it all Together"
author: "Lex Comber"
date: "June 2016"
output: pdf_document
---

# Data set up

As before, you will need to load the data from he `github` resource
```{r eval=T,  message=F, warning=F}
library(GISTools)
library(spgwr)
library(repmis)
source_data("http://github.com/lexcomber/LexTrainingR/blob/master/SpatialAccuracy.RData?raw=True")
```
And have a look at the what you have
```{r eval=T}
ls()
head(data.frame(data))
```

# Putting it all together with Loops and Functions

So far we have examined overall accuracy and per class User and Producer accuracies individually, showing the original *aspatial* or *global* measure, and then a summary of the distribution of the related *geographically weighted* values. The accuracy surfaces have then been mapped using the `level.plot` function in the `GISTools` package.

The code above has been devloped and described step by step to walk you through the process.

It might be useful to develop functions to automate some of these operations. And then perhaps to combine some of these functions so that for any given class, a number of accuracy measures are returned. The code in this section starts to do this.

# Overall Accuracy

Remember that Overall Accuracy is calculated from the sum of the diagonals in the accuracy / confusion / error / validation matrix that compares *Predicted* against *Observed* classes, divided by the total number of data points.

```{r eval=T}
res <- vector(length = dim(data)[1])
for (i in 1: dim(data)[1]) {
	if (data$Boolean_RS[i] == data$Boolean_FS[i]) {
		res[i] <- 1
	}}
```
This can be calculated from the data directly:
```{r eval=T}
cat("overall accuracy:", sum(res)/length(res))
```
Or from a logistic regression and a the `alogit` function:
```{r eval=T}
mod.ov <- glm(res~1,family= binomial) 
mod.coefs <- mod.ov$coefficients
mod.coefs[2] <-sum(mod.coefs) 
alogit <- function(x){exp(x)/(1+exp(x))}
mod.ov <- alogit(mod.coefs[2]) 
cat("overall accuracy:", mod.ov)
```

And if a `SpatialPointsDataFrame` object is created this can be used in a GW approach with the `ggwr` function. First some variables and paramters need to be set: 
```{r eval=T, results='hide', warning=F, message=F}
bw = 0.15
grid <- SpatialPoints(expand.grid(x=seq(295000,363000,by=1000),
  y=seq(3610000,3646000,by=1000)))
res.spdf <- SpatialPointsDataFrame(coords = data[,2:3], 
  data = data.frame(res))
```
Then the GW model can be constructed:
```{r eval=T, results='hide', warning=F, message=F}
gwr.mod <- ggwr(res~1, data = res.spdf, adapt = bw, 
  fit.points = grid, family= binomial) 
gwr.ov <- alogit(data.frame(gwr.mod$SDF)[,2])
```
And the variation in the distribution of Overall Accuracy values examined: 
```{r eval=T, message = F}
summary(gwr.ov)
```

## Create a Function
A function to do all of this can be assembeled from the above code snippets.

First of all it would be useful to create a `SpatialPointsDataFrame` of the all of the orginal data - this is the kind of dataset that you might bring as `shapefile` to this kind of analysis: 
```{r}
spdf <- SpatialPointsDataFrame(coords = data[,2:3], data = data.frame(data))
head(data.frame(spdf))
```
Then define a function that takes this `spdf` as input and returns a `SpatialGridDataFrame` with the results of the geographically weighted analysis:

```{r eval=T, results='hide', warning=F, message=F}
gw.accuracy <- function(spdf, Field.class = "Boolean_FS", 
    RS.class = "Boolean_RS", bw = 0.15, grid=grid, family= binomial){
  # compare predicted and observed (classified and field)
  res <- as.vector(spdf@data[RS.class] == spdf@data[Field.class]) * 1
  # notice how the line of code above replaces the specification of 
  # the  res vector, the for loop etc 
  # Commented Out: A-spatial overall accuracy
  # cat("Overall accuracy:",sum(res)/length(res), "\n")
  # GW approach
  alogit <- function(x){exp(x)/(1+exp(x))}
  gwr.mod <- ggwr(res~1, data = spdf, adapt = bw, 
    fit.points = grid, family= binomial) 
  gwr.ov <- alogit(data.frame(gwr.mod$SDF)[,2])
  # Commented Out: Summary of the GW variation
  # cat("GW overall accuracy:", summary(gwr.ov))
  # create SpatialPixelsDF to return from the function
  gw.spdf <-SpatialPixelsDataFrame(gwr.mod$SDF, data.frame(gwr.ov))
  return(gw.spdf)
}
```
And then this function can be run:  
```{r eval=T, results='hide', warning=F, message=F}
tmp <- gw.accuracy(spdf, Field.class = "Boolean_FS", 
    RS.class = "Boolean_RS", bw = 0.15, grid, family= binomial)
```
And the results in `tmp` can be evaluated:
```{r}
summary(tmp)
```

## Create a Mapping Function
And the `SpatialPixelsDataFrame` can be mapped by defining a mapping function:
```{r}
gw.mapping <- function(grd, index = 1, cols=brewer.pal(4,'Reds'), 
    bounding.poly = lib, x = 297000, y = 3650000, tit = "My Title") {
  z = data.frame(grd)[, index]
  zz = z[!is.na(z)]
  shades <- auto.shading(zz, cols = cols)
  level.plot(grd, index = index, shades = shades)
  masker = poly.outer(grd, bounding.poly, extend = 100)
  add.masking(masker) 
  plot(bounding.poly, add = T)
  choro.legend(x, y,shades) 
  title(tit)}
```
***Hint*** the `locator` function can be used to identify the `x` and `y` coordinates for the `choro.legend` function.

```{r}
par(mar = c(0,0,1,0))
gw.mapping(tmp)
```

So now there are two functions, with a number of default parameters that the user can change,  that can be called to cauluate and then map overall accuracy: 
```{r eval=F}
tmp <- gw.accuracy(spdf)
gw.mapping(tmp)
```

# User and Producer Accuracies

So far we have just mapped User and Producer accuracies for the class of `Grazing Land`. This class was chosen to expemplify the GW methods because it *does* exhibit spatial varation int hese accuracies. But this might be the case for all classes. So, in this section we will first construct a function to compute the the GW User and Producer accuracies for all classes. Then we will examine the spatial variation and select per-class accuracies to map.   

The stages in this are:

1. For each class, construct a variable pair of remote sensing (*Predcited*) class and field (*Observed*) class:
2. Compute the GW User accuracy 
3. Compute the GW Prodcuer accuracy 
4. Add the results to a data frame
5. Create a SpatialPixelsDataFrame
6. Evaluate the variation of the GW accuracy measures

```{r eval=T,  message=F, warning=F}
# define the classes
class.list <- unique(data$Boolean_RS)[order(unique(data$Boolean_RS))]
# pass this into a loop
for (i in 1:length(class.list) ){
  class <- class.list[i]	
  # 1. Construct the variable pair
  # RS indicates the class
  rs.class <- (data$Boolean_RS == class) * 1
  # FS indicates the class
  fs.class <- (data$Boolean_FS == class) * 1
  # join together
  fsrs <- data.frame(cbind(fs.class,rs.class)) 
  # convert to SPDF
  fsrs.spdf <- SpatialPointsDataFrame(coords = data[,2:3], 
    data = data.frame(fsrs))
  # 2. GW User accuracy 
  # define a bandwidth
  bw = 0.15	
  # construct GW model
  gwr.mod <- ggwr(fs.class~rs.class, data = fsrs.spdf, 
    adapt = bw,fit.points=grid, family= binomial) 
  coefs <- data.frame(gwr.mod$SDF)[,2:3]
  coefs[,2] <- rowSums(coefs) 
  alogit <- function(x){exp(x)/(1+exp(x))}
  gwr.user <- alogit(coefs[,2]) 
  # 3. GW Producer accuracy
  gwr.mod <- ggwr(rs.class~fs.class, data = fsrs.spdf, 
    adapt = bw,fit.points=grid, family= binomial) 
  coefs <- data.frame(gwr.mod$SDF)[,2:3]
  coefs[,2] <- rowSums(coefs) 
  gwr.producer <- alogit(coefs[,2]) 
  # 4. Add these to the data frame
  # define some variable names
  tit.user <- sprintf("%s-User", class)
  tit.producer <- sprintf("%s-Producer", class)
  df <- data.frame(gwr.user, gwr.producer)
  # name the df
  names(df) <- c(tit.user, tit.producer)
  # and combine
  if(i ==1) df.res <- df
  if(i > 1) df.res <- data.frame(df.res, df)
}
```
The spatial variation in the coefficients is indicated by the distribution of User accuracy values:
```{r eval=T}
summary(df.res)
```

And these can used to consruct a `SpatialPixelsDataFrame` object:
```{r eval=T}
gw.all.spdf <-SpatialPixelsDataFrame(gwr.mod$SDF, data.frame(df.res))
```

## Create a function 
This can be wrapped up into a function that takes the  `spdf` variable created above
```{r eval=T}
spdf <- SpatialPointsDataFrame(coords = data[,2:3], data = data.frame(data))
```
and returns a `SpatialPixelsDataFrame` object:
```{r eval=T,  message=F, warning=F}
user.prod.accuracy <- function(spdf, Field.class = "Boolean_FS", 
    RS.class = "Boolean_RS", bw = 0.15, grid=grid, family= binomial){
  class.list <- unique(spdf@data[,RS.class])[order(unique(spdf@data[,RS.class]))]
  # pass this into a loop
  for (i in 1:length(class.list) ){
    class <- class.list[i]	
    # 1. Construct the variable pair
    # RS indicates the class
    rs.class <- (data$Boolean_RS == class) * 1
    # FS indicates the class
    fs.class <- (data$Boolean_FS == class) * 1
    # join together
    fsrs <- data.frame(cbind(fs.class,rs.class)) 
    # convert to SPDF
    fsrs.spdf <- SpatialPointsDataFrame(coords = data[,2:3], 
      data = data.frame(fsrs))
    # 2. GW User accuracy 
    # define a bandwidth
    bw = 0.15	
    # construct GW model
    gwr.mod <- ggwr(fs.class~rs.class, data = fsrs.spdf, 
      adapt = bw,fit.points=grid, family= binomial) 
    coefs <- data.frame(gwr.mod$SDF)[,2:3]
    coefs[,2] <- rowSums(coefs) 
    alogit <- function(x){exp(x)/(1+exp(x))}
    gwr.user <- alogit(coefs[,2]) 
    # 3. GW Producer accuracy
    gwr.mod <- ggwr(rs.class~fs.class, data = fsrs.spdf, 
      adapt = bw,fit.points=grid, family= binomial) 
    coefs <- data.frame(gwr.mod$SDF)[,2:3]
    coefs[,2] <- rowSums(coefs) 
    gwr.producer <- alogit(coefs[,2]) 
    # 4. Add these to the data frame
    # define some variable names
    tit.user <- sprintf("%s-User", class)
    tit.producer <- sprintf("%s-Producer", class)
    df <- data.frame(gwr.user, gwr.producer)
    # name the df
    names(df) <- c(tit.user, tit.producer)
    # and combine
    if(i ==1) df.res <- df
    if(i > 1) df.res <- data.frame(df.res, df)
  }
  gw.spdf <-SpatialPixelsDataFrame(gwr.mod$SDF, data.frame(df.res))
  return(gw.spdf)
}
```
And then this can be called
```{r eval=F,  message=F, warning=F}
gwr.all.spdf <- user.prod.accuracy(spdf, Field.class = "Boolean_FS", 
    RS.class = "Boolean_RS", bw = 0.15, grid=grid, family= binomial)
```
Or accepting the defaults:
```{r eval=T,  message=F, warning=F}
gwr.all.spdf <- user.prod.accuracy(spdf)
```
And the contents examined again: 
```{r eval=T}
summary(gwr.all.spdf@data)
```

The elements of the `gwr.all.spdf` variable can be mapped using the fucntion defined earlier: 
```{r eval=F}
par(mar = c(0,0,1,0))
gw.mapping(gwr.all.spdf, tit = names(gwr.all.spdf)[1])
```
Of course the parameters can be adjusted:
```{r eval=F}
gw.mapping(gwr.all.spdf, index = 2, tit = names(gwr.all.spdf)[2])
gw.mapping(gwr.all.spdf, index = 5, tit = names(gwr.all.spdf)[5], 
  cols = brewer.pal(6,'Spectral'))
# or write to a file
png(filename = "plot.png")
gw.mapping(gwr.all.spdf, index = 5, tit = names(gwr.all.spdf)[5], 
  cols = brewer.pal(6,'YlOrRd'))
dev.off()
```
Or even put into a loop:
```{r eval=F}
for (i in seq(2, 10, by = 2)) {
  gw.mapping(gwr.all.spdf, index = i, tit = names(gwr.all.spdf)[i])
}  
```
And other shading schemes are available - see:
```{r eval=F}
display.brewer.all()
```

